---
title: "COMPMUT Dynamics 1: Generation of dynamics figures"
date: "compiled Jan 2024"
author: "jpjh"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(tidyverse)
library(ggplot2)
library(knitr)
source("./functions/theme_pub.r")

theme_github <- function(x){
  theme_pub(base_size=12)
}
theme_set(theme_github())
```

### Read in and pre-process data

Plots for the dynamics experiments.

Read in data frame.

```{r}
df <- read.csv("./data/2_FractionData.csv", header=TRUE)

summary(df) %>% kable()
```

Summarise variables for examination.

Rough plot of everything to see how things look.

```{r}
df %>% pivot_longer(starts_with("Fraction"), names_to = "Subpop", values_to = "Fraction") %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Subpop)) +
  geom_area() + 
  facet_grid(Treatment ~ Replicate + Ratio) +
  scale_fill_viridis_d()
```

Looks good.

Plot each experiment in turn as described in the manuscript.

Experiment 1 part 1: Effect of selection.

Treatments under consideration:

- A SBW25.GmR ; SBW25.GmR pQBR57d59.GFP ; SBW25d4242.GmR.dTomato pQBR57
- AM SBW25.GmR ; SBW25.GmR pQBR57d59.GFP ; SBW25d4242.GmR.dTomato pQBR57 + Mercury
- C SBW25.GmR ; SBW25.GmR pQBR57d59.tdTomato ; SBW25d4242.GmR.YFP pQBR57
- CM SBW25.GmR ; SBW25.GmR pQBR57d59.tdTomato ; SBW25d4242.GmR.YFP pQBR57 + Mercury

For A and AM, chrCM = Fraction_Yellow and plaCM = Fraction_Red. For C and CM, chrCM = Fraction_Red and plaCM = Fraction_Yellow.

```{r}
(df_expt1 <- df %>%
  filter(Experiment == 1 & Treatment %in% c("A", "AM", "C", "CM")) %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Population = factor(
    case_when(startsWith(Treatment, "A") & Fluorescence == "Yellow" ~ "plaCM",
              startsWith(Treatment, "A") & Fluorescence == "Red" ~ "chrCM",
              startsWith(Treatment, "C") & Fluorescence == "Yellow" ~ "chrCM",
              startsWith(Treatment, "C") & Fluorescence == "Red" ~ "plaCM",
              Fluorescence == "Orange" ~ "double / clumped",
              Fluorescence == "Blue" ~ "none"),
    levels=c("double / clumped", "chrCM", "plaCM", "none")),
         Selection = ifelse(endsWith(Treatment, "M"), "Yes", "No"),
         Markers = ifelse(startsWith(Treatment, "A"), "Red / Yellow",
                          "Yellow / Red")))

(p1_supp <- df_expt1 %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Selection + Markers ~ Replicate) +
  scale_fill_manual(
    breaks=c("chrCM", "plaCM", "none", "double / clumped"),
    values=c("#882255", "#44AA99", "#DDDDDD", "#000000")))
```

Looks good — now make an image with averaging over reps.

```{r}
(p1 <- df_expt1 %>%
  group_by(Population, Markers, Selection, Transfer) %>%
  summarise(mean = mean(Fraction), 
            n = n(), 
            se = sd(Fraction)/sqrt(n), 
            ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
  rename(Fraction=mean) %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Selection ~ Markers) +
  scale_fill_manual(
    breaks=c("chrCM", "plaCM", "none", "double / clumped"),
    values=c("#882255", "#44AA99", "#DDDDDD", "#000000")))
```

To create the main figure for the paper, also average over markers, and copy the t=0 data to the Selection=Yes treatment.

```{r}
df_expt1_summ <- df_expt1 %>%
  group_by(Population, Selection, Transfer) %>%
  summarise(mean = mean(Fraction), 
            n = n(), 
            se = sd(Fraction)/sqrt(n), 
            ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
  rename(Fraction=mean)

(p1_summ <- df_expt1_summ %>% filter(Transfer==0) %>%
  mutate(Selection = "Yes") %>%
  bind_rows(df_expt1_summ) %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
    ggtitle("Mercury selection?") +
  facet_grid(. ~ Selection) +
  scale_fill_manual(
    breaks=c("chrCM", "plaCM", "none", "double / clumped"),
    values=c("#882255", "#44AA99", "#DDDDDD", "#000000")))

png("./figs/p1.png", width=3.6, height=2.4, units="in", res=300)
p1_summ + theme_pub() + theme(legend.position="bottom")
dev.off()

tiff("./figs/Fig3.tiff", width=3.6, height=2.4, units="in", res=300)
p1_summ + theme_pub() + theme(legend.position="bottom")
dev.off()

png("./figs/p1_supp.png", width=7.2, height=4, units="in", res=300)
p1_supp + theme_pub() + theme(legend.position="bottom")
dev.off()
```

Now experiment 2.

```{r}
(df_expt2 <- df %>%
  filter(Experiment == 1 & Treatment %in% c("B", "BM", "D", "DM")) %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Population = factor(
    case_when(startsWith(Treatment, "B") & Fluorescence == "Yellow" ~ "plaCM",
              startsWith(Treatment, "B") & Fluorescence == "Red" ~ "chrCM",
              startsWith(Treatment, "D") & Fluorescence == "Yellow" ~ "chrCM",
              startsWith(Treatment, "D") & Fluorescence == "Red" ~ "plaCM",
              Fluorescence == "Orange" ~ "double / clumped",
              Fluorescence == "Blue" ~ "pQBR103 only"),
    levels=c("double / clumped", "chrCM", "plaCM", "pQBR103 only")),
         Selection = ifelse(endsWith(Treatment, "M"), "Yes", "No"),
         Markers = ifelse(startsWith(Treatment, "B"), "Red / Yellow",
                          "Yellow / Red")))

(p2_supp <- df_expt2 %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Selection + Markers ~ Replicate) +
  scale_fill_manual(
    breaks=c("chrCM", "plaCM", "pQBR103 only", "double / clumped"),
    values=c("#882255", "#44AA99", "#004488", "#000000")))
```

And the summary plot:

```{r}
(p2 <- df_expt2 %>%
  group_by(Population, Markers, Selection, Transfer) %>%
  summarise(mean = mean(Fraction), 
            n = n(), 
            se = sd(Fraction)/sqrt(n), 
            ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
  rename(Fraction=mean) %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Selection ~ Markers) +
  scale_fill_manual(
    breaks=c("chrCM", "plaCM", "pQBR103 only", "double / clumped"),
    values=c("#882255", "#44AA99", "#004488", "#000000")))
```

And the average:

```{r}
df_expt2_summ <- df_expt2 %>%
  group_by(Population, Selection, Transfer) %>%
  summarise(mean = mean(Fraction), 
            n = n(), 
            se = sd(Fraction)/sqrt(n), 
            ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
  rename(Fraction=mean)

(p2_summ <- df_expt2_summ %>% filter(Transfer==0) %>%
  mutate(Selection = "Yes") %>%
  bind_rows(df_expt2_summ) %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(. ~ Selection) +
    ggtitle("Mercury selection?") +
  scale_fill_manual(
    breaks=c("chrCM", "plaCM", "pQBR103 only", "double / clumped"),
    values=c("#882255", "#44AA99", "#004488", "#000000")))

png("./figs/p2.png", width=3.6, height=2.4, units="in", res=300)
p2_summ + theme_pub() + theme(legend.position="bottom")
dev.off()

tiff("./figs/Fig4.tiff", width=3.6, height=2.4, units="in", res=300)
p2_summ + theme_pub() + theme(legend.position="bottom")
dev.off()

png("./figs/p2_supp.png", width=7.2, height=4, units="in", res=300)
p2_supp + theme_pub() + theme(legend.position="bottom")
dev.off()
```

Plot the controls.

```{r}
(df_ctrls <- df %>% filter(startsWith(Treatment, "F")) %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Controls = case_when(Treatment == "F1" ~ "Yellow plaCM only",
                                Treatment == "F2" ~ "Red chrCM only",
                                Treatment == "F3" ~ "Red plaCM only",
                                Treatment == "F4" ~ "Yellow chrCM only"))) %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Fluorescence)) +
  geom_area() + 
  facet_grid(Controls ~ Replicate) +
  scale_fill_viridis_d()
```

### The host availability experiments

First head-to-head CMs.

```{r}
(df_expt3 <- df %>%
  filter(Experiment == 2 & Treatment == "A") %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Population = factor(
    case_when(Fluorescence == "Yellow" ~ "plaCM",
              Fluorescence == "Red" ~ "chrCM",
              Fluorescence == "Orange" ~ "double / clumped",
              Fluorescence == "Blue" ~ "none"),
    levels=c("double / clumped", "chrCM", "plaCM", "none")),
         Ratio = factor(Ratio,
                        levels=c(10, 1, 0.1, 0),
                        labels=c("10:1", "1:1", "1:10", "none"))))

(p3_supp <- df_expt3 %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Replicate ~ Ratio) +
    scale_fill_manual(
    breaks=c("chrCM", "plaCM", "none", "double / clumped"),
    values=c("#882255", "#44AA99", "#DDDDDD", "#000000")))
```

And the summary figure:

```{r}
(p3 <- df_expt3 %>%
   group_by(Population, Ratio, Transfer) %>%
   summarise(mean = mean(Fraction), 
             n = n(), 
             se = sd(Fraction)/sqrt(n), 
             ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
   rename(Fraction=mean) %>%
   ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
   geom_area() + 
   ggtitle("Recipient availability") +
   facet_grid(. ~ Ratio) +
   scale_fill_manual(
     breaks=c("chrCM", "plaCM", "none", "double / clumped"),
     values=c("#882255", "#44AA99", "#DDDDDD", "#000000")))

png("./figs/p3.png", width=4.6, height=2.4, units="in", res=300)
p3 + theme_pub() + theme(legend.position="bottom")
dev.off()

tiff("./figs/Fig5.tiff", width=4.6, height=2.4, units="in", res=300)
p3 + theme_pub() + theme(legend.position="bottom")
dev.off()

png("./figs/p3_supp.png", width=4.6, height=7.2, units="in", res=300)
p3_supp + theme_pub() + theme(legend.position="bottom")
dev.off()
```

Then the plasmids from the corresponding CMs:

```{r}
(df_expt4 <- df %>%
  filter(Experiment == 2 & Treatment == "B") %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Population = factor(
    case_when(Fluorescence == "Yellow" ~ "plaCM",
              Fluorescence == "Red" ~ "pQBR57 (wild-type) starting in chrCM",
              Fluorescence == "Orange" ~ "double / clumped",
              Fluorescence == "Blue" ~ "none"),
    levels=c("double / clumped", "pQBR57 (wild-type) starting in chrCM", "plaCM", "none")),
         Ratio = factor(Ratio,
                        levels=c(10, 1, 0.1, 0),
                        labels=c("10:1", "1:1", "1:10", "none")))) 

(p4_supp <- df_expt4 %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Replicate ~ Ratio) +
    scale_fill_manual(
    breaks=c("pQBR57 (wild-type) starting in chrCM", "plaCM", "none", "double / clumped"),
    values=c("#CC6677", "#44AA99", "#DDDDDD", "#000000")))
```

And the summary figure:

```{r}
(p4 <- df_expt4 %>%
   group_by(Population, Ratio, Transfer) %>%
   summarise(mean = mean(Fraction), 
             n = n(), 
             se = sd(Fraction)/sqrt(n), 
             ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
   rename(Fraction=mean) %>%
   ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
   geom_area() + 
   ggtitle("Recipient availability") +
   facet_grid(. ~ Ratio) +
    scale_fill_manual(
    breaks=c("pQBR57 (wild-type) starting in chrCM", "plaCM", "none", "double / clumped"),
    values=c("#CC6677", "#44AA99", "#DDDDDD", "#000000")))

png("./figs/p4.png", width=4.6, height=2.4, units="in", res=300)
p4 + theme_pub() + theme(legend.position="bottom")
dev.off()

tiff("./figs/Fig6.tiff", width=4.6, height=2.4, units="in", res=300)
p4 + theme_pub() + theme(legend.position="bottom")
dev.off()

png("./figs/p4_supp.png", width=4.6, height=7.2, units="in", res=300)
p4_supp + theme_pub() + theme(legend.position="bottom")
dev.off()
```

And finally the wild-type vs. plaCM

```{r}
(df_expt5 <- df %>%
  filter(Experiment == 2 & Treatment == "C") %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Population = factor(
    case_when(Fluorescence == "Yellow" ~ "plaCM",
              Fluorescence == "Red" ~ "pQBR57 (wild-type) starting in uncompensated",
              Fluorescence == "Orange" ~ "double / clumped",
              Fluorescence == "Blue" ~ "none"),
    levels=c("double / clumped", "pQBR57 (wild-type) starting in uncompensated", "plaCM", "none")),
         Ratio = factor(Ratio,
                        levels=c(10, 1, 0.1, 0),
                        labels=c("10:1", "1:1", "1:10", "none")))) 

(p5_supp <- df_expt5 %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Replicate ~ Ratio) +
    scale_fill_manual(
    breaks=c("pQBR57 (wild-type) starting in uncompensated", "plaCM", "none", "double / clumped"),
    values=c("#DDCC77", "#44AA99", "#DDDDDD", "#000000")))
```

And the summary figure:

```{r}
(p5 <- df_expt5 %>%
   group_by(Population, Ratio, Transfer) %>%
   summarise(mean = mean(Fraction), 
             n = n(), 
             se = sd(Fraction)/sqrt(n), 
             ci = (qt(0.95/2 + 0.5, n-1)) * se) %>%
   rename(Fraction=mean) %>%
   ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
   geom_area() + 
   ggtitle("Recipient availability") +
   facet_grid(. ~ Ratio) +
    scale_fill_manual(
    breaks=c("pQBR57 (wild-type) starting in uncompensated", "plaCM", "none", "double / clumped"),
    values=c("#DDCC77", "#44AA99", "#DDDDDD", "#000000")))

png("./figs/p5.png", width=4.6, height=2.4, units="in", res=300)
p5 + theme_pub() + theme(legend.position="bottom")
dev.off()

tiff("./figs/Fig7.tiff", width=4.6, height=2.4, units="in", res=300)
p5 + theme_pub() + theme(legend.position="bottom")
dev.off()

png("./figs/p5_supp.png", width=4.6, height=7.2, units="in", res=300)
p5_supp + theme_pub() + theme(legend.position="bottom")
dev.off()
```

There was also an experiment performed with *P. putida* KT2440 as an alternative recipient.

```{r}
(df_expt5 <- df %>%
  filter(Experiment == 2 & Treatment == "D") %>%
  pivot_longer(starts_with("Fraction"), names_to = "Fluorescence", names_prefix = "Fraction_",
               values_to = "Fraction") %>%
  mutate(Population = factor(
    case_when(Fluorescence == "Yellow" ~ "plaCM",
              Fluorescence == "Red" ~ "chrCM",
              Fluorescence == "Orange" ~ "double / clumped",
              Fluorescence == "Blue" ~ "KT2440"),
    levels=c("double / clumped", "chrCM", "plaCM", "KT2440")),
         Ratio = factor(Ratio,
                        levels=c(10, 1, 0.1, 0),
                        labels=c("10:1", "1:1", "1:10", "none")))) %>%
  ggplot(aes(x=Transfer, y=Fraction, fill=Population)) +
  geom_area() + 
  facet_grid(Replicate ~ Ratio) +
    scale_fill_manual(
    breaks=c("chrCM", "plaCM", "KT2440", "double / clumped"),
    values=c("#882255", "#44AA99", "#332288", "#000000")) +
  theme(legend.position="bottom")
```

However, in all cases *P. putida* outcompeted *P. fluorescens* SBW25, with no difference between the CMs. As these experiments did not reveal anything of interest regarding plasmid biology they were not included in the manuscript. 

---

**[Back to index.](../README.md)**
