---
title: "COMPMUT Dynamics 3: Estimation and gathering of relevant parameters"
date: "compiled November 2023"
author: "jpjh"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(tidyverse)
library(ggplot2)
library(deSolve)
library(knitr)
source("./functions/theme_pub.r")
theme_github <- function(x){
  theme_pub(base_size=12)
}
theme_set(theme_github())
```

## Background

#### Growth rates

Estimate growth rates from a plate-reader experiment, using the R package `gcplyr` and values identified in a preliminary analysis. Measurement cycles every 15 minutes. 

```{r}
library(gcplyr)

read.csv("./data/4_WTGrowthCurve.csv", header=TRUE) %>%
  group_by(well, replicate) %>%
  mutate(Time = cycle / 4,
         smoothed = smooth_data(x = Time, y = value,
                                sm_method = "moving-average", window_width_n = 9),
         smderivpercap_3 = calc_deriv(x = Time, y = smoothed,
                                       percapita = TRUE, blank = 0, window_width_n = 3)) %>%
  summarize(max_percap = max_gc(smderivpercap_3, na.rm = TRUE),
            max_percap_time = extr_val(Time, which_max_gc(smderivpercap_3)),
            doub_time = doubling_time(y = max_percap)) %>% 
  ungroup() %>%
  summarise(mean = mean(max_percap),
            sd = sd(max_percap)) %>% kable()
```

Results in an estimate:

- alpha = 0.54 ± 0.06.

Generate a table of values.

```{r}
parameters <- data.frame(
  parameter = c("alpha"),
  value = c(0.5412246),
  sd = c(0.0553524)
)
```

This value is consistent with other unpublished measurements for SBW25 wild-type growth rate in the plate reader.

#### Plasmid fitness costs

The actual mean fitness values for the chromosomal and plasmid mutations, from [Hall et al. 2021 PLoS Biology](dx.doi.org/10.1371/journal.pbio.3001225) (see also www.github.com/jpjh/COMPMUT), were:

```{r}
source("https://raw.githubusercontent.com/jpjh/COMPMUT/main/functions/calculate_fitness.R")
id_cols <- c("experiment","replicate","host","plasmid")

chromosomal_compensation_marker <- calculate_fitness(
  read.csv("https://raw.githubusercontent.com/jpjh/COMPMUT/main/data/COMPMUT_exp_data_1.csv", header=TRUE), id_cols) %>%
  filter(host=="wild-type" & plasmid=="plasmid-free") %>% 
  summarise(mean = mean(W_gm)) %>% pull()    

(chromosomal_compensation_fitness <- calculate_fitness(
  read.csv("https://raw.githubusercontent.com/jpjh/COMPMUT/main/data/COMPMUT_exp_data_1.csv", header=TRUE), id_cols) %>% 
    filter(host %in% c("wild-type", "delta_PFLU4242") &
                      plasmid %in% c("plasmid-free", "pQBR57")) %>%
    mutate(W = W_gm/chromosomal_compensation_marker) %>%
  group_by(host, plasmid) %>% summarise(fitness = mean(W), sd = sd(W)))

id_colsp <- c("replicate","marker","plasmid")

plasmid_compensation_marker <- calculate_fitness(
  read.csv("https://raw.githubusercontent.com/jpjh/COMPMUT/main/data/COMPMUT_exp_data_2.csv", header=TRUE), id_colsp) %>%
  filter(plasmid=="plasmid-free") %>% 
  summarise(mean = mean(W_gm)) %>% pull()    

(plasmid_compensation_fitness <- calculate_fitness(
  read.csv("https://raw.githubusercontent.com/jpjh/COMPMUT/main/data/COMPMUT_exp_data_2.csv", header=TRUE), id_colsp) %>%
    mutate(W = ifelse(marker == "Sm", 1/(W_gm / plasmid_compensation_marker),
                               (W_gm / plasmid_compensation_marker)),
           host = "wild-type") %>%
  group_by(host, plasmid) %>% summarise(fitness = mean(W), sd = sd(W)))

(compensations <- bind_rows(chromosomal_compensation_fitness, plasmid_compensation_fitness) %>%
  filter(plasmid!="plasmid-free") %>%
  bind_cols(data.frame(strain = c("chromosome compensation","no compensation","plasmid compensation","no compensation")))) %>%
  kable()
```

i.e.

- beta has two estimates: 0.82 ± 0.05; 0.81 ± 0.03
- beta_chrCM = 0.97 ± 0.07
- beta_plaCM = 0.94 ± 0.04

```{r}
betas <- data.frame(
  parameter = c("beta","beta_chrCM","beta_plaCM"),
  value = c(mean(c(0.8238168,0.8122002)), 0.9690664, 0.9449607),
  sd = c(mean(c(0.0520151,0.0337998)), 0.0698719, 0.0413181)
)

parameters <- rbind(parameters, betas)
```


#### Carrying capacity

Taken from the endpoint of the plasmid-free competitions in the PLoS Biology paper.

```{r}
read.csv("https://raw.githubusercontent.com/jpjh/COMPMUT/main/data/COMPMUT_exp_data_1.csv", header=TRUE) %>% 
  filter(timepoint=="end" & plasmid=="plasmid-free" & host=="wild-type") %>%
  summarise(cfu_ml = mean((count_white + count_blue) * (1000/spread) * 10^dilution)/1e6,
            sd = sd((count_white + count_blue) * (1000/spread) * 10^dilution)/1e6) %>% kable()
```

- K = 5.7e9 ± 1.6e9

```{r}
parameters[5,] <- c("K",5.71625e9,1.590994e9)
```


#### Gamma

Taken from the PLoS Biology paper, with estimate from the uncompensated strains.

```{r}
d7 <- read.csv("https://raw.githubusercontent.com/jpjh/COMPMUT/main/data/COMPMUT_exp_data_3-1.csv", header=TRUE)

d7_start_D <- d7 %>% 
  filter(timepoint == "start_donor") %>% select(!recipient & !timepoint) %>%
  mutate(D0 = (10^dilution) * (1000/spread) * count_donor * 0.5 * 0.01)

d7_start_R_pf <- d7 %>%
  filter(timepoint == "start_pf_recipient") %>% select(!donor & !plasmid & !timepoint) %>%
  mutate(R0 = (10^dilution) * (1000/spread) * count_recipient * 0.5 * 0.01)

d7_start_pf <- full_join(d7_start_D, d7_start_R_pf, by=c("replicate"),
                         suffix=c("donor","recipient")) %>%
  select(replicate, donor, plasmid, recipient, D0, R0)

d7_end_DR <- d7 %>% filter(timepoint == "end_DR") %>%
  mutate(D = (10^dilution) * (1000/spread) * count_donor,
         R = (10^dilution) * (1000/spread) * count_recipient) %>%
  select(replicate, donor, plasmid, recipient, D, R)

d7_end_T <- d7 %>% filter(timepoint == "end_T") %>%
  mutate(TC = (10^dilution) * (1000/spread) * count_recipient) %>%
  select(replicate, donor, plasmid, recipient, TC)

left_join(d7_start_pf, d7_end_DR) %>%
  left_join(d7_end_T) %>%
  mutate(N0 = D0 + R0,
         N = D + R,
         gamma = (log(N/N0)/24) * log(1+((TC/R)*(N/D))) * (1/(N-N0))) %>%
  filter(donor == "wild-type" & plasmid == "pQBR57_anc") %>%
  summarise(mean = mean(gamma),
            sd = sd(gamma),
            log10mean = mean(log10(gamma)),
            log10sd = sd(log10(gamma))) %>% kable(digits=20)
```

Estimates:

- gamma = 10^-11.35±0.1 i.e. 4.5e-12 ± 1e-12

```{r}
parameters[6,] <- c("gamma",4.573199e-12,1.013806e-12)
```


#### Mu

Mu is the washout/death rate. 

Populations were transferred every 24h losing 99% of their population each transfer.

Therefore estimate mu at:

```{r}
(mu <- 0.99/24)

parameters[7,] <- c("mu", mu, 0)
```

### Table of parameters

```{r}
parameters %>% kable()

write.table(parameters, file = c("./data/3_Parameters.txt"),
            sep="\t", quote=FALSE, row.names=FALSE)
```

---

**[Back to index.](../README.md)**
